---
title: "Hands_on_sessions_5_EHR"
author: "MARKO LUDAIC"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"      
output:
  html_document:
    toc: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


# Exercise 1

You will download a subset of EHR data from the campus global. The data contains only the hospital admission data, patient information and diagnosis data. 

```{r}
library(data.table)
codes <- fread("data/allDiseaseCodes.txt", quote = "")
diagnosis <- fread("data/diagnosisData.txt", quote = "")
admission <- fread("data/admissionData.txt", quote = "")
patient <- fread("data/patientData.txt", quote = "")

```


# EDA

Perform an EDA for the data. 
Some guide questions for the EDA are:

## top diseases in the dataset

What are the most prevalent diseases?

```{r}
# using to display the number of each disease's diagnosis
table(diagnosis$diagnosis_code)
# sort by decreasing order to see what are the top ones
sort(table(diagnosis$diagnosis_code), decreasing = TRUE)
# retrieve the names of diseases for the top 10 diagnosis
# list of diagnosis codes you want to filter for
diagnosis_codes <- c("4019", "2724", "53081", "25000", "311", "42731", "V1582", "4280", "41401", "3051")
# Filter the rows that contain the diagnosis codes of interest and select the diagnosis_code and diagnosis_description columns
top_10_diseases <- unique(subset(diagnosis, diagnosis_code %in% diagnosis_codes, select = c(diagnosis_code, diagnosis_description)))
# Convert diagnosis_code to a factor with the levels ordered according to the diagnosis_codes vector
top_10_diseases$diagnosis_code <- factor(top_10_diseases$diagnosis_code, levels = diagnosis_codes)
# Use arrange function from dplyr package to sort the rows based on diagnosis_code
library(dplyr)
top_10_diseases <- arrange(top_10_diseases, diagnosis_code)
# Display the resulting table
top_10_diseases
```
- The top 3 most prevalent diseases taking into account all patients, both male and female, are "Unspecified essential hypertension", "Other and unspecified hyperlipidemia" and "Esophageal reflux". Top 10 prevalent diseases are represented in the table, as the diagnosis' codes and description. 

### top diseases by gender

Are they different is data is stratified by gender? 

```{r}
# Subset patient data into separate data frames for males and females
males <- filter(patient, patient_sex == "M")
females <- filter(patient, patient_sex == "F")

# Filter diagnosis data by patient IDs in male and female data frames
diagnosis_males <- filter(diagnosis, patient_id %in% males$patient_id)
diagnosis_females <- filter(diagnosis, patient_id %in% females$patient_id)


### MALES
# sort by decreasing order to see what are the top ones
sort(table(diagnosis_males$diagnosis_code), decreasing = TRUE)
male_codes <- c("4019", "2724", "25000", "53081", "41401", "V1582", "42731", "4280", "3051", "5849")
top_10_diseases_male <- unique(subset(diagnosis_males, diagnosis_code %in% male_codes, select = c(diagnosis_code, diagnosis_description)))
# Convert diagnosis_code to a factor with the levels ordered according to the diagnosis_codes vector
top_10_diseases_male$diagnosis_code <- factor(top_10_diseases_male$diagnosis_code, levels = male_codes)
# Use arrange function from dplyr package to sort the rows based on diagnosis_code
top_10_diseases_male <- arrange(top_10_diseases_male, diagnosis_code)
# Display the resulting table
top_10_diseases_male


### FEMALES
# sort by decreasing order to see what are the top ones
sort(table(diagnosis_females$diagnosis_code), decreasing = TRUE)
female_codes <- c("4019", "2724", "53081", "311", "2449", "25000","4280", "42731", "V1582", "30000")
top_10_diseases_female <- unique(subset(diagnosis_females, diagnosis_code %in% female_codes, select = c(diagnosis_code, diagnosis_description)))
# Convert diagnosis_code to a factor with the levels ordered according to the diagnosis_codes vector
top_10_diseases_female$diagnosis_code <- factor(top_10_diseases_female$diagnosis_code, levels = female_codes)
# Use arrange function from dplyr package to sort the rows based on diagnosis_code
top_10_diseases_female <- arrange(top_10_diseases_female, diagnosis_code)
# Display the resulting table
top_10_diseases_female
# combining tables to visualize all together
combined_sex <- cbind(top_10_diseases_female, top_10_diseases_male)
colnames(combined_sex)<-c("diagnosis_code_female","description_female","diagnosis_code_male","description_male")
combined_sex
```



# VISUALIZATION

```{r}
library(ggplot2)


### FEMALES
# Sort the table in descending order and select only the top 10 elements
top_10_diagnoses_female <- head(sort(table(diagnosis_females$diagnosis_code), decreasing = TRUE), 10)

# Create a data frame with the top 10 diagnoses and their counts
df <- data.frame(count = as.numeric(top_10_diagnoses_female),
                 name = top_10_diseases_female$diagnosis_description)

# Create a histogram with ggplot2
females_hist<-ggplot(df, aes(x = name, y = count)) +
  geom_bar(stat = "identity", fill = "blue") +
  xlab("Diagnosis") +
  ylab("Count") +
  ggtitle("Top 10 Diagnoses in Females") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1),
        plot.title = element_text(size = 10))


### MALES
# Sort the table in descending order and select only the top 10 elements
top_10_diagnoses_male <- head(sort(table(diagnosis_males$diagnosis_code), decreasing = TRUE), 10)

# Create a data frame with the top 10 diagnoses and their counts
df <- data.frame(count = as.numeric(top_10_diagnoses_male),
                 name = top_10_diseases_male$diagnosis_description)


# Create a histogram with ggplot2
males_hist<-ggplot(df, aes(x = name, y = count)) +
  geom_bar(stat = "identity", fill = "blue") +
  xlab("Diagnosis") +
  ylab("Count") +
  ggtitle("Top 10 Diagnoses in Males") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1),
        plot.title = element_text(size = 10))


library(gridExtra)
grid.arrange(females_hist, males_hist, ncol = 2)
```

- As it can be seen from the graphical representation of male and female's top 10 diagnoses, "Unspecified essential hypertention" seems to be the top one diagnosis for both sex followed by "Other and unspecified hyperlipidemia" as a top 2nd diagnosis for both sex. However, there are some variations regarding top 10 diagnosis when stratifies by sex. Females are more prone to diagnoses such as "Esophageal reflux" and Unspecified acquired hypothyrodism". Males are more prone to "Other and unspecified hyperlipidemia".

What are the most prevalent diseases when aggregating codes to the level of chapters?

```{r}
# preserve only first 3 numbers of the diagnosis_code, for the categorization
diagnosis$col <- substr(diagnosis$diagnosis_code, 1, 3)
disease_codes <- diagnosis$col
chapters <- c("001-139", "140-239", "240-279", "280-289", "290-319", "320-389", "390-459","460-519", "520-579", "580-629", "630-679", "680-709", "710-739", "740-759", "760-779", "780-799", "800-999")

diagnosis$chapter<-NA

map_to_chapter <- function(code) {
  # Check if the code is in any of the chapter ranges
  for (i in seq_along(chapters)) {
    chapter_range <- strsplit(chapters[i], "-")[[1]]
    if (length(chapter_range) == 1) {
      if (code == chapter_range) {
        return(chapters[i])
      }
    } else {
      if (code >= chapter_range[1] & code <= chapter_range[2]) {
        return(chapters[i])
      }
    }
  }
  # Check for E chapter
  if (substring(code, 1, 1) == "E") {
    return("E000-E999")
  }
  # Check for V chapter
  if (substring(code, 1, 1) == "V") {
    return("V01-V91")
  }
  return(NA)
}

# Create a new column with the chapter for each code in col
diagnosis$chapter <- sapply(diagnosis$col, map_to_chapter)
```


```{r}
# assigning descriptions to chapter codes
# descriptions have been found on the ICD website
lookup_table <- data.frame(
  chapter = c("001-139", "140-239", "240-279", "280-289", "290-319", "320-389", "390-459",
              "460-519", "520-579", "580-629", "630-679", "680-709", "710-739", "740-759",
              "760-779", "780-799", "800-999", "V01-V91", "E000-E999"),
  chapter_description = c("Infectious And Parasitic Diseases", "Neoplasms", "Endocrine, Nutritional And Metabolic Diseases, And Immunity Disorders",
                           "Diseases Of The Blood And Blood-Forming Organs", "Mental Disorders", "Diseases Of The Nervous System And Sense Organs",
                           "Diseases Of The Circulatory System", "Diseases Of The Respiratory System", "Diseases Of The Digestive System",
                           "Diseases Of The Genitourinary System", "Complications Of Pregnancy, Childbirth, And The Puerperium",
                           "Diseases Of The Skin And Subcutaneous Tissue", "Diseases Of The Musculoskeletal System And Connective Tissue",
                           "Congenital Anomalies", "Certain Conditions Originating In The Perinatal Period", "Symptoms, Signs, And Ill-Defined Conditions",
                           "Injury And Poisoning", "Supplementary Classification Of Factors Influencing Health Status And Contact With Health Services",
                           "Supplementary Classification Of External Causes Of Injury And Poisoning")
)

# I will assign it to a new data frame diagnosis_chap (chapters)
diagnosis_chap <- merge(diagnosis, lookup_table, by = "chapter", all.x = TRUE)

```


```{r}
# sort by decreasing order to see what are the top ones
sort(table(diagnosis_chap$chapter), decreasing = TRUE)

# list of diagnosis codes to filter for
chap_codes <- c("390-459", "240-279", "290-319", "520-579", "320-389", "580-629", "710-739")

# Convert diagnosis_code to a factor with the levels ordered according to the chap_codes vector
diagnosis_chap$chapter <- factor(diagnosis_chap$chapter, levels = chap_codes)

# sorting the table by the chapter column
top_10_chapters <- diagnosis_chap[diagnosis_chap$chapter %in% chap_codes,]
top_10_chapters <- top_10_chapters[order(top_10_chapters$chapter),]

# removing duplicates
top_10_chapters <- unique(top_10_chapters[, c("chapter", "chapter_description")])

# the resulting table
top_10_chapters

```
- When aggregating codes to the level of chapters the most prevalent diseases are diseases of the circulatory system, endocrine, nutritional and metabolic diseases with immunity disorders and mental disorders. I excluded three sections that are not considered diagnosis, but if taken into account, they would be in the top 10 most prevalent "diseases". These include "V01-V91" which group a wide range of conditions, such as abnormal findings on examination of blood and urine, symptoms involving respiratory, digestive, cardiovascular or nervous systems, and ill-defined conditions, such as malaise and fatigue. The next two are "E000-E999" (additional information about the circumstances surrounding an injury or illness) and "780-799" (describing symptoms and abnormal findings, used to provide more complete picture of patient's condition).

# Comorbidities of Obesity

What are the diseases that co-occur with obesity?   
Hint: use the field “Description” in the codes file to retrieve the ICD 9 codes containing the string “obesity”  
Alternatively, use the corresponding ICD 9 code. Important note: for these data, the dot of the ICD-9 code has been removed 
```{r}
# pick up all observations that mention "obesty" in the long_title column
obesity <- codes[grepl("obesity", codes$long_title, ignore.case = TRUE), ]
# extract their codes
obesity_codes<-obesity$icd_code
# find out which patients have obesity diagnosis
patients <- diagnosis[diagnosis$diagnosis_code %in% obesity_codes & !duplicated(diagnosis$patient_id), ]
# extract patients' IDs
obesity_patients <- patients$patient_id
# merge data frames 
comorbidities <- diagnosis[diagnosis$patient_id %in% obesity_patients, ]
```

- I made new data frame called "comorbidities" which contains only, and all patients that have obesity, amongst other diagnosis. So, by searching for the top prevalent diseases, it can be found out what diseases co-occur with obesity. 

```{r}
# sort by decreasing order to see what are the top ones
sort(table(comorbidities$diagnosis_code), decreasing = TRUE)
comorb_codes <- c("27800", "4019", "27801", "53081", "49390", "32723", "311", "V270", "30000", "3051")
top_10_obesity <- unique(subset(comorbidities, diagnosis_code %in% comorb_codes, select = c(diagnosis_code, diagnosis_description)))
# Convert diagnosis_code to a factor with the levels ordered according to the diagnosis_codes vector
top_10_obesity$diagnosis_code <- factor(top_10_obesity$diagnosis_code, levels = comorb_codes)
# Use arrange function from dplyr package to sort the rows based on diagnosis_code
top_10_obesity <- arrange(top_10_obesity, diagnosis_code)
# Display the resulting table
top_10_obesity
```

- It seems that the top 1 disease that co-occur with obesity is "Unspecified essential hypertension" which is then followed by "Esophageal reflux" and "Asthma, unspecified type, unspecified" as the next two comorbidities.


```{r}
################## using the ELIXHAUSER METHOD #####################

install.packages("comorbidity")
library(comorbidity)

obes<-comorbidity(x = diagnosis, id = "patient_id", code = "diagnosis_code", map = "elixhauser_icd9_quan", assign0 = FALSE)

obes_comorb<-obes[obes$obes==1,]

# find columns that contain 1

cols_with_1 <- colnames(obes_comorb)[apply(obes_comorb, 2, function(x) any(x == 1))]


# create a table of counts
count_table <- matrix(0, nrow = 1, ncol = length(cols_with_1), dimnames = list(c("Count"), cols_with_1))
for (i in 1:length(cols_with_1)) {
  count_table[1, i] <- sum(obes_comorb[[cols_with_1[i]]] == 1)
}

# sort the table in descending order
count_table <- count_table[, order(-count_table)]

# Print the sorted table
count_table

```

- Based on these results and the scores that have been obtained, "Hypertension,uncomplicated", "Diabetes,uncomplicated" and "Chronic pulmonary disease" are the top three comorbidities of obesity. 

## Comorbidities of Obesity stratified by gender

Are the same comorbidities occurring in men and in women ?  


Present the results in a publication-like manner: appropriately use tables and/or different types of plots
Are there any of the comorbidities supported by the literature? Do they have common genetic basis?


```{r}
### FEMALES
comorbidities_female <- diagnosis_females[diagnosis_females$patient_id %in% obesity_patients, ]
sort(table(comorbidities_female$diagnosis_code), decreasing = TRUE)
comorb_codes_female <- c("27800", "4019", "27801", "53081", "49390", "V270", "311", "64911", "30000", "32723")
top_10_obesity_female <- unique(subset(comorbidities_female, diagnosis_code %in% comorb_codes_female, select = c(diagnosis_code, diagnosis_description)))
# Convert diagnosis_code to a factor with the levels ordered according to the diagnosis_codes vector
top_10_obesity_female$diagnosis_code <- factor(top_10_obesity_female$diagnosis_code, levels = comorb_codes_female)
# Use arrange function from dplyr package to sort the rows based on diagnosis_code
top_10_obesity_female <- arrange(top_10_obesity_female, diagnosis_code)
# Display the resulting table
top_10_obesity_female

###############

### MALES
comorbidities_male <- diagnosis_males[diagnosis_males$patient_id %in% obesity_patients, ]
sort(table(comorbidities_male$diagnosis_code), decreasing = TRUE)
comorb_codes_male <- c("27800", "4019", "32723", "27801", "53081", "3051", "311", "V1582", "30000", "42731")
top_10_obesity_male <- unique(subset(comorbidities_male, diagnosis_code %in% comorb_codes_male, select = c(diagnosis_code, diagnosis_description)))
# Convert diagnosis_code to a factor with the levels ordered according to the diagnosis_codes vector
top_10_obesity_male$diagnosis_code <- factor(top_10_obesity_male$diagnosis_code, levels = comorb_codes_male)
# Use arrange function from dplyr package to sort the rows based on diagnosis_code
top_10_obesity_male <- arrange(top_10_obesity_male, diagnosis_code)
# Display the resulting table
top_10_obesity_male

###############

# combining tables to visualize all together
combined_comorb <- cbind(top_10_obesity_female, top_10_obesity_male)
colnames(combined_comorb)<-c("diagnosis_code_female","description_female","diagnosis_code_male","description_male")
combined_comorb
```

- When stratifying by gender it seems that, more or less, there are the same comorbidities occuring with obesity, but with different frequency. Additionally, there are some small differences in terms of some disease being present as a comorbidity in males, but not in females, and vice versa. "Unspecified essential hypertension" seems to be the first disease that co-occures with obesity in both males and females, follower by "Esophageal reflux". "Tobacco use disorder" seems to affect only males, while symptomes regarding pregnancy and childbirth follow obesity in females. 

# session info {.unnumbered}

```{r, results='asis',  echo=FALSE, message=FALSE }
sessionInfo()
```
